---
title: "Discovery_comparison"
output: BiocStyle::html_document
date: "06/05/2024"
author: "Jana Braunger"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
BiocStyle::html_document 

data_dir = '/Users/janabraunger/Documents/Projects/scCRISPR/data/'
figdir = '/Users/janabraunger/Documents/Projects/scCRISPR/figures/discovery_comparison/'
if(!dir.exists(figdir)) dir.create(figdir)
knitr::opts_chunk$set(dev = c("png", "pdf"), fig.path = figdir)
```

# Required libraries and data
## Load needed libraries
```{r load_libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(GGally)
library(pheatmap)
```

## Load guide assignment results
```{r}
# data frame containing the cells with single assignments per method
guide_assignment <- read_csv(paste0(data_dir, 'guide_calling/Replogle_K562/', 
                                    'single_assignments_all_methods.csv'), show_col_types = FALSE) 
head(guide_assignment)

n_assigned_cells <- guide_assignment %>% 
  group_by(method) %>%
  summarize(n_cells = n())
head(n_assigned_cells)
```

```{r}
# create a coloring df
colors <- c('UMI_t30' = "#828282", 'Maximum' = "#9ace25", 'ratio_50%' = "#009500", 
           'Cell Ranger' = "#06cdff", 'Poisson-Gauss' = "#06A4FF", 
           '2-Beta' = "#e6c202", '3-Beta' = "#e68902",
           'Poisson' = "#e63102", 'Negative Binomial' = "#e60291", 
           'Binomial' = "#d102e6", 'top_5%_cells' = "#8102e6")
write_csv(data.frame(colors) %>% rownames_to_column(var = 'method'), 
          paste0(data_dir, 'sceptre_pipeline/Replogle_K562/colors.csv'))
```


## Load power check results for various guide calling approaches
```{r}
get_power_check_results <- function(data, name, method_names){
  method_name = method_names[method_names$dir_name == name, 'method']
  df = readRDS(paste0(data_dir, 'sceptre_pipeline/', data, '/power_check/', name, 
                                '/results_run_power_check.rds')) %>%
    filter(pass_qc == TRUE) %>%
    mutate(method = method_name)
  df <- df %>% mutate(p_adj = p.adjust(df$p_value, method="BH"))
  return(df)
}
```

```{r}
# define all methods
dir_names = c("2-BetaMM", "3-BetaMM", "cellranger", "replogle", 
              "SCEPTRE", "negative_binomial", "maximum", "binomial", 
              "quantiles/q0.05", "UMI/t30", "ratios/t0.5")
method_names <- data.frame('method' = c("2-Beta", "3-Beta", "Cell Ranger", "Poisson-Gauss", 
                                        "Poisson", "Negative Binomial", "Maximum", "Binomial",
                                        "top_5%_cells", "UMI_t30", "ratio_50%"), 
                           'dir_name' = dir_names)
data = 'Replogle_K562'

power_check <- lapply(dir_names, function(name) {get_power_check_results(data, name, method_names)}) %>% 
  bind_rows() %>%
  mutate(method = factor(method, levels = names(colors), ordered = TRUE))
head(power_check)
```


## Load calibration results for various guide calling approaches
```{r}
get_calibration_results <- function(data, name, method_names){
  method_name = method_names[method_names$dir_name == name, 'method']
  df = readRDS(paste0(data_dir, 'sceptre_pipeline/', data, '/discovery_analysis/', name, 
                           '/results_run_calibration_check.rds')) %>%
    filter(pass_qc == TRUE) %>%
    mutate(method = method_name)
  df <- df %>% mutate(p_adj = p.adjust(df$p_value, method="BH"))
  return(df)
}
```

```{r}
calibration <- lapply(dir_names, function(name) {get_calibration_results(data, name, method_names)}) %>% 
  bind_rows() %>%
  group_by(method) %>%
  mutate(total_tests = n()) %>%
  filter(significant == TRUE) %>%
  group_by(method, total_tests) %>%
  summarize(false_discoveries = n()) %>% 
  right_join(method_names) %>% 
  replace(is.na(.), 0) %>%
  mutate(method = factor(method, levels = names(colors), ordered = TRUE))
```


## Load discovery analysis results for various guide calling approaches
```{r}
get_discovery_analysis_results <- function(data, name, method_names){
  method_name = method_names[method_names$dir_name == name, 'method']
  df <- readRDS(paste0(data_dir, 'sceptre_pipeline/', data, '/discovery_analysis/', name, 
                           '/results_run_discovery_analysis.rds')) %>%
    filter(pass_qc == TRUE, significant == TRUE) %>%
    mutate(method = method_name)
}
```

```{r}
discovery_analysis <- lapply(dir_names, function(name) {get_discovery_analysis_results(data, name, method_names)}) %>% 
  bind_rows() %>%
  mutate(method = factor(method, levels = names(colors), ordered = TRUE))
head(discovery_analysis)
```


# Analysis of true positives (power check on target genes)
```{r sig_downreg_targets, fig.height=3.2, fig.width = 4.4}
# Number of significantly downregulated target genes
signif_targets <- power_check %>% filter(p_adj < 0.05) %>% 
  group_by(method) %>% 
  summarize(signif_targets = n()) %>%
  left_join(n_assigned_cells)

ggplot(signif_targets, aes(x = n_cells, y = signif_targets, group = method)) +
  geom_point(aes(color = method), size = 2) + 
  theme_bw() + 
  labs(x = 'Number of assigned cells', y = 'Significant target genes') +
  scale_color_manual(name = 'Guide assignment', values = colors, 
                     breaks = names(colors)) 
```


```{r pairwise_log2fc, fig.height=10, fig.width=10, warning=FALSE}
# log2 fold changes of the target genes
lfc <- power_check %>% 
  mutate(neg_lfc = - log_2_fold_change) %>%
  subset(select = c(grna_target, neg_lfc, method)) %>% 
  spread(key = method, value = neg_lfc) %>%
  column_to_rownames(var = "grna_target") %>% 
  rename("Negative\nBinomial" = "Negative Binomial") 
dim(lfc[complete.cases(lfc), ])

lowerFn <- function(data, mapping) {
  p <- ggplot(data = data, mapping = mapping) +
    geom_point(size = 0.3, alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")
  p
}

ggpairs(lfc,
  diag = list(continuous = "density"),
  upper = list(continuous = "cor"),
  lower = list(continuous = wrap(lowerFn)),
  axisLabels = "show") +
  theme_bw()

lfc <- lfc %>% rename("Negative Binomial" = "Negative\nBinomial") 
```


```{r difference_heatmap_targets, fig.width=3.8, fig.height=3.3}
# calculate pairwise differences
calculate_diff <- function(lfc, method1, method2){
  diff <- (lfc[[method2]] - lfc[[method1]]) / sqrt(2)
  return(median(diff, na.rm = TRUE))
}

# create data frame with pairwise methods
pairs = expand.grid(colnames(lfc), colnames(lfc))
pairs$diff <- apply(pairs, 1, function(row) calculate_diff(lfc, row[1], row[2]))
diff_matrix <- pairs %>% 
  spread(key = Var2, value = diff) %>% 
  column_to_rownames(var = "Var1")
diff_matrix <- diff_matrix[names(colors), names(colors)]

# plot heatmap
color_breaks <- seq(-0.8, 0.8, length.out = 100)
pheatmap(diff_matrix, display_numbers = F, 
         cluster_rows = F, cluster_cols = F, scale = "none",
         color = rev(colorRampPalette((RColorBrewer::brewer.pal(n = 7, name ="RdBu")))(100)),
         breaks = color_breaks,
         legend_title = "Median difference of log2 fold changes", 
         gaps_row = c(1, 3, 5), gaps_col = c(1, 3, 5))
```

```{r median_lfc, fig.height=3.5, fig.width = 5}
# median log2 fold change over number of assigned cells
medians <- apply(lfc, 2, function(x) median(x, na.rm = TRUE))
med_lfc <- data.frame(median_lfc = medians) %>% 
  rownames_to_column(var = "method") %>% 
  left_join(n_assigned_cells)

ggplot(med_lfc, aes(x = n_cells, y = median_lfc, group = method)) + 
  geom_point(aes(color = method), size = 2) + 
  theme_bw() + 
  labs(x = 'Number of assigned cells', y = 'Median negative log2 fold change') +
  scale_color_manual(name = 'Guide assignment', values = colors, 
                     breaks = names(colors))
```


# Discovery analysis 
```{r hist_n_response_genes}
# Get the number of response genes per perturbation
n_discovery <- discovery_analysis %>% group_by(grna_target, method) %>%
  summarize(n_response_genes = n())

ggplot(n_discovery, aes(x = n_response_genes)) +
  geom_histogram(bins = 50) + 
  facet_wrap(~method) +
  theme_bw() +
  xlab('Number of response genes per perturbation') +
  ylab('Counts') 
```

```{r median_total_response_genes, fig.width=12, fig.height = 3}
# plot median and sum of response genes per method
stats <- n_discovery %>% 
  group_by(method) %>% 
  summarize(median = median(n_response_genes), total = sum(n_response_genes)) %>% 
  gather(key = metric, value = value, -method)

ggplot(stats, aes(x = method, y = value, fill = method)) + 
  geom_bar(stat = 'identity') +
  facet_wrap(~metric, scales = 'free_y') +
  theme_bw() + 
  theme(legend.position = "none") +
  xlab("") +
  ylab("")
```


```{r scatter_false_total_discoveries, fig.width=4.4, fig.height=3.3}
# scatter plot with number of total discoveries vs number of assigned cells
discoveries <- stats %>% 
  filter(metric == 'total') %>% 
  rename(total_discoveries = value) %>% 
  subset(select = -metric) %>% 
  left_join(n_assigned_cells)

ggplot(discoveries, aes(x = n_cells, y = total_discoveries, group = method)) +
  geom_point(aes(color = method), size = 2) +
  theme_bw() + 
  labs(color = 'Guide calling method', x = 'Number of assigned cells', y = 'Number of total discoveries') +
  scale_color_manual(name = 'Guide assignment', values = colors, 
                     breaks = names(colors)) #+
  #scale_shape_manual(name = 'Guide assignment', values = shapes)
```

```{r prop_false_positives, fig.width=11, fig.height=2.5}
ggplot(calibration, aes(x = method, y = false_discoveries, fill = method)) +
  geom_bar(stat = 'identity') + 
  theme_bw() + 
  labs(x = 'Method', y = 'Number of false discoveries') + 
  theme(legend.position = "none") +
  scale_fill_manual(values = colors)
```

# Data set summary 
## Quality comparison
```{r Replogle_summary_heatmap, fig.width=3, fig.height=4}
results <- results <- power_check %>% 
  group_by(method) %>% 
  summarize(neg_lfc = - median(log_2_fold_change, na.rm = TRUE)) %>%
  left_join(signif_targets %>% select(method, signif_targets)) %>%
  left_join(discoveries %>% select(method, total_discoveries)) %>%
  column_to_rownames(var = "method")
results <- as.data.frame(scale(results))

seq_breaks <- c(seq(min(results), 0, (0-min(results))/100 * 2), seq(0, max(results), (max(results)-0)/100 * 2)[-1])
pheatmap(results, labels_col = c("Target downregulation", "Significant targets",
                                 "Total discoveries"),
         cluster_rows=FALSE, cluster_cols = FALSE, breaks = seq_breaks, gaps_row = c(1, 3, 5),
         color = rev(colorRampPalette((RColorBrewer::brewer.pal(n = 7, name ="RdBu")))(100)))
```


## Run time comparison
```{r}
get_run_time <- function(file_dir, name, method_names){
  method_name = method_names[method_names$dir_name == name, 'method']
  df = read_csv(file_dir, show_col_types = F) 

  # Estimate full time for methods that were run in parallel for gRNA sets
  if (method_name %in% c('Poisson', 'Negative Binomial')){
    df <- mutate(df, time = time / 200 * 2282)
  }
  if (method_name %in% c('Binomial')){
    df <- mutate(df, time = time / 400 * 2282)
  }
  df <- df %>%
    mutate(method = method_name) %>%
    mutate(mins = time / 60) %>%
    mutate(h = mins / 60)
  return(df)
}
```

```{r gc_run_time, fig.width = 8, fig.height = 2.5}
# Guide calling
dir_names_gc = c("2-BetaMM", "3-BetaMM", "cellranger", "replogle", 
              "SCEPTRE", "negative_binomial", "maximum", "binomial",
              "quantiles", "UMI", "ratios")
method_names_gc <- data.frame('method' = c("2-Beta", "3-Beta", "Cell Ranger", "Poisson-Gauss", 
                                        "Poisson", "Negative Binomial", "Maximum", "Binomial",
                                        "top_5%_cells", "UMI_t30", "ratio_50%"), 
                           'dir_name' = dir_names_gc) 

gc_times <- lapply(dir_names_gc, function(name) {
  get_run_time(paste0(data_dir, 'guide_calling/', data, '/', name, '/run_time.csv'), 
               name, method_names_gc)}) %>% 
  bind_rows() %>%
  mutate(method = if_else(method == "Negative Binomial", "Negative\nBinomial", method)) %>%
  mutate(method = if_else(method == "top_5%_cells", "Top 5%\ncells", method)) %>%
  mutate(method = if_else(method == "Poisson-Gauss", "Poisson-\nGauss", method)) 

names(colors)[names(colors) == "Negative Binomial"] <- "Negative\nBinomial"
names(colors)[names(colors) == "top_5%_cells"] <- "Top 5%\ncells"
names(colors)[names(colors) == "Poisson-Gauss"] <- "Poisson-\nGauss"

ggplot(gc_times, aes(x = reorder(method, h), y = h, color = method)) + 
  geom_point(size = 2) +
  theme_bw() + 
  labs(x = '', y = 'Run time [h]', title = 'Guide assignment') + 
  theme(legend.position = "none") +
  scale_color_manual(values = colors)
```

```{r}
get_run_time <- function(file_dir, name, method_names){
  method_name = method_names[method_names$dir_name == name, 'method']
  df = read_csv(file_dir, show_col_types = F) 
  df <- df %>%
    mutate(method = method_name) %>%
    mutate(mins = time / 60) %>%
    mutate(h = mins / 60)
  return(df)
}
```


```{r power_check_run_time, fig.width = 8, fig.height = 2.5}
power_times <- lapply(dir_names, function(name) {
  get_run_time(paste0(data_dir, 'sceptre_pipeline/', data, '/power_check/', name, '/run_times.csv'), 
               name, method_names)}) %>% 
  bind_rows() %>%
  filter(step == 'power check') %>%
  mutate(method = if_else(method == "Negative Binomial", "Negative\nBinomial", method)) %>%
  mutate(method = if_else(method == "top_5%_cells", "Top 5%\ncells", method)) %>%
  mutate(method = if_else(method == "Poisson-Gauss", "Poisson-\nGauss", method))

ggplot(power_times, aes(x = reorder(method, h), y = mins, color = method)) + 
  geom_point(size = 2) +
  theme_bw() + 
  labs(x = '', y = 'Run time [mins]', title = 'Power check')+ 
  theme(legend.position = "none") +
  scale_color_manual(values = colors)
```

```{r calibration_run_time, fig.width = 8, fig.height = 2.5}
discovery <- lapply(dir_names, function(name) {
  get_run_time(paste0(data_dir, 'sceptre_pipeline/', data, '/discovery_analysis/', name, '/run_times.csv'), 
               name, method_names)}) %>% 
  bind_rows()  %>%
  mutate(method = if_else(method == "Negative Binomial", "Negative\nBinomial", method)) %>%
  mutate(method = if_else(method == "top_5%_cells", "Top 5%\ncells", method)) %>%
  mutate(method = if_else(method == "Poisson-Gauss", "Poisson-\nGauss", method))

ggplot(filter(discovery, step == 'calibration check'), aes(x = reorder(method, h), y = h, color = method)) + 
  geom_point(size = 2) +
  theme_bw() + 
  labs(x = '', y = 'Run time [h]', title = 'Calibration check')+ 
  theme(legend.position = "none") +
  scale_color_manual(values = colors)
```

```{r discovery_run_time, fig.width = 8, fig.height = 2.5}
ggplot(filter(discovery, step == 'discovery analysis'), aes(x = reorder(method, h), y = h, color = method)) + 
  geom_point(size = 2) +
  theme_bw() + 
  labs(x = '', y = 'Run time [h]', title = 'Discovery Analysis')+ 
  theme(legend.position = "none") +
  scale_color_manual(values = colors) 
```


```{r summary_run_time, fig.width=3, fig.height=4}
names(colors)[names(colors) == "Negative\nBinomial"] <- "Negative Binomial"
names(colors)[names(colors) == "Top 5%\ncells"] <- "top_5%_cells"
names(colors)[names(colors) == "Poisson-\nGauss"] <- "Poisson-Gauss"

# Run time summary heatmap
run_times <- gc_times %>% subset(select = c(method, h)) %>%
  rename(gc_time = h) %>%
  left_join(subset(discovery, select = c(method, step, h)) %>% filter(step %in% c('discovery analysis')) %>% spread(step, h)) %>%
  mutate(method = if_else(method == "Negative\nBinomial", "Negative Binomial", method)) %>%
  mutate(method = if_else(method == "Top 5%\ncells", "top_5%_cells", method)) %>%
  mutate(method = if_else(method == "Poisson-\nGauss", "Poisson-Gauss", method)) %>%
  left_join(n_assigned_cells) %>%
  mutate(method = factor(method, levels = names(colors), ordered = TRUE)) %>%
  arrange(method) %>%
  column_to_rownames(var = 'method') 
run_times <- as.data.frame(scale(run_times))

seq_breaks <- c(seq(min(run_times), 0, (0-min(run_times))/100 * 2), seq(0, max(run_times), (max(run_times)-0)/100 * 2)[-1])
pheatmap(run_times, labels_col = c("Guide calling time", "Discovery time", "Number of cells"),
         cluster_rows=FALSE, cluster_cols = FALSE, breaks = seq_breaks, gaps_row = c(1, 3, 5),
         color = rev(colorRampPalette((RColorBrewer::brewer.pal(n = 7, name ="PRGn")))(100)))
```


# Session info
```{r}
sessionInfo()
```

